Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-20 sources/bmc_template_submission/bmc_article.tex
MK edits verification test

We pruned 9 out of the 41 original tips of the Gottlieb et al. tree, corresponding
to a 20\% trim, excluding the outgroups. We then performed a Physcraper run to test
if we would recover the dropped tips.
We successfully recovered 6 out of 9 pruned tips in the updated tree.
Closer examination of results revealed that the 3 missing tips were correctly retrieved
with BLAST along with the other 6 tips that were included in the updated tree. Examining
the ITS sequences belonging to the tips missed by the test, by following the GenBank
accession numbers reported in the original publication, showed that these sequences
contain a 100 bp long gap of unidentified nucleotides (Ns). These Ns were absent
in the 3 sequences published in the original alignment.

@@

Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.
The three sequences that were in the original tree but were not included in
the updated tree by Physcraper in the verification test were excluded because
they contain a 100 bp long gap of unidentified nucleotides (Ns) in GenBank that
was manually curated and removed from in the original alignment.

These sequences do appear in the Physcraper results in the list of matches from GenBank which did not fit the sequence length cutoffs set in the configuration.
This ``seqlen\_mismatch.txt" file includes the accession number, taxon, and sequence length of all sequences filtered based on sequence length.

>>>>

To test the accuracy of Physcraper we designed a verification test.
We pruned 9 out of the 41 original tips of the Gottlieb2005 tree \cite{gottlieb2005molecular},
corresponding to a 20\% trim, excluding the outgroups. We then performed a Physcraper
run to test if we would recover the dropped tips.
We successfully recovered 6 out of 9 pruned tips in the updated tree.
Closer examination of results revealed that sequences for the 3 missing tips were correctly retrieved
with BLAST along with sequences from the other 6 tips recovered in the updated tree.
By following the GenBank accession numbers reported in the original publication
belonging to the ITS sequences of the missing tips, we observed that these three sequences
contain a 100 bp long gap of unidentified nucleotides (Ns) that is completely absent
from any of the sequences in the original alignment. This caused these three GenBank sequences
in particular to exceed Physcraper's default sequence length cutoff of 120\%, being thus
filtered and excluded from the alignment step onwards.
These sequences do appear in the Physcraper results in the list of matches from
GenBank which did not fit the sequence length cutoffs set in the configuration file.
This ``seqlen\_mismatch.txt" file includes the accession number, taxon, and sequence
length of all sequences filtered based on sequence length.

@@

Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.
